---
title: "EDUCE descriptive publication data figures"
author: "Kim Dill-McFarland"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
# Setup
## Load packages
```{r message=FALSE}
#Data manipulation and figures
library(tidyverse)
#Multi-panel figures
library(cowplot)
#Exact and Monte Carlo symmetry tests for paired contigency tables
library(rcompanion)
```

## R session
```{r}
sessionInfo()
```

# Community of practice
## Load data
```{r}
cop <- read_tsv("data_clean/2017.18.19_EDUCEteam.txt")
cop
```

Calculate totals by career level (*e.g.* student, faculty, etc.) in each Department
```{r}
cop_sum <- cop %>% 
  #Gather career levels into 1 column
  gather(key="key", value="total", -dept, -fac) %>% 
  #Remove NAs
  filter(!is.na(total)) %>% 
  #Sum totals of each career level in each department
  group_by(key, dept, fac) %>% 
  summarize(n=sum(total)) %>% 
  #Reorder variables for plot
  mutate(key_ord = factor(key, 
                          levels=c("Undergraduate","Graduate",
                                   "Postdoc","Instructor","Research",
                                   "Staff"))) %>% 
  mutate(dept_ord = factor(dept,
                    levels=c("IAM","BOTA","Central","CPSC","ECE","LFS","MATH","MGEN","MICB","STAT"))) %>% 
  mutate(fac_ord = factor(fac, levels=c("ASCI","LFS","MED","SCIE","Central")))
```

## Figure 3. EDUCE team members at UBC
```{r}
cop_plot <- cop_sum %>%
  #Create variable for Science vs Other facets
  mutate(fac_group = ifelse(fac == "SCIE", fac, "Other")) %>% 
  #Reorder variable for facets
  mutate(fac_group_ord = factor(fac_group, levels=c("SCIE","Other"))) %>% 
  
#Plot
  ggplot() +
  geom_bar(aes(x=key_ord, y=n, fill=dept_ord), stat="identity", width = 0.5) +
  facet_grid(~fac_group_ord, scales = "free_x", space="free",
             labeller = as_labeller(c("SCIE"="Faculty of Science",
                          "Other"="Other"))) +
  #Beautify
  theme_classic() +
  labs(x="Career level", y="Number of EDUCE\nteam members") +
  theme(legend.key.height = unit(0.75, "cm"),
        text=element_text(colour="black", size=10)) +
  scale_x_discrete(labels=c("Staff" = "Staff",
                            "Research"="Faculty\n(research)",
                            "Instructor" = "Faculty\n(instructor)",
                            "Postdoc" = "Postdoctoral\nfellow",
                            "Graduate" = "Graduate\nstudent",
                            "Undergraduate" = "Undergraduate\nstudent")) +
  scale_fill_brewer(name = "Department", labels = c("Applied Mathematics",
                                                    "Botany",
                                                    "Centre for Teaching,\nLearning & Technology",
                                                    "Computer Science",
                                                    "Electrical & Computer\nEngineering",
                                                    "Food Science",
                                                    "Mathematics",
                                                    "Medical Genetics",
                                                    "Microbiology &\nImmunology",
                                                    "Statistics"),
                    palette = "RdBu") +
  scale_y_continuous(breaks=c(0:6))

cop_plot
```

Save Figure 3
```{r}
ggsave(filename="Fig3.pdf", plot=cop_plot, width=19.05, height=9, units = "cm")
```

\pagebreak

# Student interest and experience
## Load data
```{r}
survey <- read_csv("data_clean/2017.18.19_survey_clean.csv")
```

*Information on survey data clean-up can be found in `EDUCE_survey_cleanup`*

## Survey question: How would you rate your **interest** in...
### Monte Carlo symmetry tests for paired contingency tables
Pre responses in columns, post responses in rows

Data cleaning
```{r}
interest <- survey %>% 
  # Select variables of interest
  select(Course, year,
         Pre_Interest_BI, Post_Interest_BI,
         Pre_Interest_CPSC, Post_Interest_CPSC,
         Pre_Interest_STAT, Post_Interest_STAT) %>% 
  # Filter to just MICB301 course data
  filter(Course == "MICB301") %>% 
  # Convert numeric survey respones to groups
  ## None=0, low=1-3, med=4-7, high=8-10
  ### Create row ID to keep matched responses together
  rowid_to_column() %>%
  gather(key="key", value="value", 
         -Course, -year, -rowid) %>% 
  mutate(value = ifelse(value %in% c("0",0), "None",
                     ifelse(value %in% c("1","2","3"), "Low",
                     ifelse(value %in% c("4","5","6","7"), "Medium",
                     ifelse(value %in% c("8","9","10"), "High",
                     value))))) %>% 
  group_by(rowid) %>% 
  spread(key=key, value=value) %>% 
  ungroup()
```

#### Bioinformatics
Test for differences in pre- vs. post- matched surveys.
```{r}
BI_interest <- interest %>% 
  # Select variables of interest
  select(Pre_Interest_BI, Post_Interest_BI) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Interest_BI, Post_Interest_BI) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Interest_BI, n) %>% 
  replace(is.na(.), 0) %>% 
  # add Pre=none
  mutate(None=c(0,0,0,0)) %>% 
  #Order variables
  select(Post_Interest_BI, None, Low, Medium, High) %>% 
  arrange(factor(Post_Interest_BI, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Interest_BI")
  
BI_interest

nominalSymmetryTest(BI_interest, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

#### Computer science
Test for differences in pre- vs. post- matched surveys.
```{r}
CPSC_interest <- interest %>% 
  # Select variables of interest
  select(Pre_Interest_CPSC, Post_Interest_CPSC) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Interest_CPSC, Post_Interest_CPSC) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Interest_CPSC, n) %>% 
  replace(is.na(.), 0) %>% 
  #Order variables
  select(Post_Interest_CPSC, None, Low, Medium, High) %>% 
  arrange(factor(Post_Interest_CPSC, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Interest_CPSC")

CPSC_interest

nominalSymmetryTest(CPSC_interest, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

#### Statistics
Test for differences in pre- vs. post- matched surveys.
```{r}
STAT_interest <- interest %>% 
  # Select variables of interest
  select(Pre_Interest_STAT, Post_Interest_STAT) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Interest_STAT, Post_Interest_STAT) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Interest_STAT, n) %>% 
  replace(is.na(.), 0) %>% 
  # Add Pre=None
  mutate(None=c(0,0,0,0)) %>% 
  #Order variables
  select(Post_Interest_STAT, None, Low, Medium, High) %>% 
  arrange(factor(Post_Interest_STAT, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Interest_STAT")

STAT_interest

nominalSymmetryTest(STAT_interest, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

### Interest plot
Data cleaning
```{r}
plot_I_dat <- interest %>% 
  #Gather pre/post data
  gather("subject", "interest", -Course, -year, -rowid) %>% 
  drop_na(interest) %>% 
  # Create separate pre/post ID column
  separate(subject, into=c("survey","trash","subject"), sep="_") %>% 
  #Reorder groups
  mutate(survey =  factor(survey, levels = c("Pre", "Post")),
         interest = factor(interest, levels=c("High","Medium","Low","None"))) %>% 
  #Remove trash column containing just "Interest" part of name
  select(-trash) %>% 
  # Calculate percentages of responses
  group_by(Course, survey, subject, interest) %>% 
  summarize(n=n()) %>% 
  mutate(freq=100*n/sum(n))
```

Bar plot
```{r}
plot_I <- ggplot(plot_I_dat, 
                 aes(x=survey, y=freq, fill=interest)) + 
  geom_col(position = "fill", width=0.5) +
  #Beautify
  labs(x="Survey", y="Proportion of\nresponses", fill="") +
  facet_grid(~subject, labeller = as_labeller(c("BI"="Bioinformatics\nP = 1.82E-2", 
                                                "CPSC"="Computer science\n",
                                                "STAT"="Statistics\n"))) +
  theme_classic(base_size = 16) +
  theme(text = element_text(size=13), 
        panel.spacing = unit(2, "lines")) +
  scale_x_discrete(labels=c("Pre","Post")) +
  scale_fill_brewer(palette = "Blues", direction=-1,
                    name="Interest") +
  scale_y_continuous(labels=scales::percent)

## Add significant arrows
arrow_I_bi<-data.frame(
  x = 1, y = 0.5, xend = 2, yend = 0.8, 
  subject=factor("BI", levels=c("BI","CPSC","STAT")))

plot_I <- plot_I + geom_segment(data=arrow_I_bi, aes(x=x, y=y, xend=xend, yend=yend), 
               arrow = arrow(length = unit(0.03, "npc")),
               inherit.aes = FALSE) 

plot_I
```

## Survey question: What level of **experience** do you have in ...
### Monte Carlo symmetry tests for paired contingency tables
Pre responses in columns, post responses in rows

Data cleaning
```{r}
exp <- survey %>% 
  # Select variables of interest
  select(Course, year,
         Pre_Exp_BI, Post_Exp_BI,
         Pre_Exp_CPSC, Post_Exp_CPSC,
         Pre_Exp_STAT, Post_Exp_STAT) %>% 
  # Filter to just MICB301 course data
  filter(Course == "MICB301") %>% 
  # Convert numeric survey respones to groups
  ## None=0, low=1-3, med=4-7, high=8-10
  ### Create row ID to keep matched responses together
  rowid_to_column() %>%
  gather(key="key", value="value", 
         -Course, -year, -rowid) %>% 
  mutate(value = ifelse(value == "0", "None",
                     ifelse(value %in% c("1","2","3"), "Low",
                     ifelse(value %in% c("4","5","6","7"), "Medium",
                     ifelse(value %in% c("8","9","10"), "High",
                     value))))) %>% 
  #  Convert 1 "very high" response to "high"
  mutate(value = ifelse(value=="veryHigh","High",value)) %>% 
  #Spread back to wide format
  group_by(rowid) %>% 
  spread(key=key, value=value) %>% 
  ungroup()
```

#### Bioinformatics
Test for differences in pre- vs. post- matched surveys.
```{r}
BI_exp <- exp %>% 
  # Select variables of exp
  select(Pre_Exp_BI, Post_Exp_BI) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Exp_BI, Post_Exp_BI) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Exp_BI, n) %>% 
  replace(is.na(.), 0) %>% 
  # Add data for Pre = High since none exist
  mutate(High = c(0,0,0,0)) %>% 
  #Order variables
  select(Post_Exp_BI, None, Low, Medium, High) %>% 
  arrange(factor(Post_Exp_BI, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Exp_BI")

BI_exp

nominalSymmetryTest(BI_exp, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

#### Computer science
Test for differences in pre- vs. post- matched surveys.
```{r}
CPSC_exp <- exp %>% 
  # Select variables of interest
  select(Pre_Exp_CPSC, Post_Exp_CPSC) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Exp_CPSC, Post_Exp_CPSC) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Exp_CPSC, n) %>% 
  replace(is.na(.), 0) %>% 
  #Order variables
  select(Post_Exp_CPSC, None, Low, Medium, High) %>% 
  arrange(factor(Post_Exp_CPSC, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Exp_CPSC")

CPSC_exp

nominalSymmetryTest(CPSC_exp, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

#### Statistics
Test for differences in pre- vs. post- matched surveys.
```{r}
STAT_exp <- exp %>% 
  # Select variables of interest
  select(Pre_Exp_STAT, Post_Exp_STAT) %>% 
  drop_na() %>% 
  # Count matched pre-post response
  group_by(Pre_Exp_STAT, Post_Exp_STAT) %>% 
  summarize(n=n()) %>% 
  # Format into contingency table
  spread(Pre_Exp_STAT, n) %>% 
  replace(is.na(.), 0) %>% 
  # Add Pre = High data
  mutate(High = c(0,0,0,0)) %>% 
  #Order variables
  select(Post_Exp_STAT, None, Low, Medium, High) %>% 
  arrange(factor(Post_Exp_STAT, levels = c("None", "Low", "Medium", "High"))) %>% 
  column_to_rownames(var="Post_Exp_STAT")

STAT_exp

nominalSymmetryTest(STAT_exp, digits=5, method="fdr", MonteCarlo=TRUE, ntrial=10000)
```

### Experience plot
Data cleaning
```{r}
plot_E_dat <- exp %>% 
  #Gather pre/post data
  gather("subject", "exp", -Course, -year, -rowid) %>% 
  drop_na() %>% 
  # Create separate pre/post ID column
  separate(subject, into=c("survey","trash","subject"), sep="_") %>% 
  #Reorder groups
  mutate(survey =  factor(survey, levels = c("Pre", "Post")),
         exp = factor(exp, levels=c("veryHigh","High","Medium","Low","None"))) %>% 
  #Remove trash column containing just "Interest" part of name
  select(-trash) %>% 
  # Calculate percentages of responses
  group_by(Course, survey, subject, exp) %>% 
  summarize(n=n()) %>% 
  mutate(freq=100*n/sum(n)) 
```

Bar plot
```{r}
plot_E <- ggplot(plot_E_dat, aes(x=survey, y=freq)) + 
  geom_col(aes(fill=exp), position = "fill", width=0.5) +
  #Beautify
  labs(x="Survey", y="Proportion of\nresponses", fill="") +
  facet_grid(~subject, labeller = as_labeller(
    c("BI"="Bioinformatics\nP < 0.03",
      "CPSC"="Computer science\nP = 2.44E-3",
      "STAT"="Statistics\n"))) +
  theme_classic(base_size = 16) +
  theme(text = element_text(size=13), 
        panel.spacing = unit(2, "lines")) +
  scale_x_discrete(labels=c("Pre","Post")) +
  scale_fill_brewer(palette = "Blues", direction=-1,
                    name="Experience") +
  scale_y_continuous(labels=scales::percent)

#Add arrows
arrow_bi<-data.frame(
  x=1,xend=2, y1=0.6,yend1=0.85, y2=0.2,yend2=0.35, y3=0.2,yend3=0.8, 
  subject=factor("BI", levels=c("BI","CPSC","MICB")))

arrow_cpsc<-data.frame(
  x=1, xend=2, y=0.2,yend=0.4,
  subject=factor("CPSC", levels=c("BI","CPSC","MICB")))

plot_E <- plot_E +
  geom_segment(data=arrow_bi, aes(x=x, y=y1, xend=xend, yend=yend1), 
               arrow = arrow(length = unit(0.03, "npc"))) +
  geom_segment(data=arrow_bi, aes(x=x, y=y2, xend=xend, yend=yend2), 
               arrow = arrow(length = unit(0.03, "npc"))) +
  geom_segment(data=arrow_bi, aes(x=x, y=y3, xend=xend, yend=yend3), 
               arrow = arrow(length = unit(0.03, "npc"))) +
  
  geom_segment(data=arrow_cpsc, aes(x=x, y=y, xend=xend, yend=yend), 
               arrow = arrow(length = unit(0.03, "npc")))

plot_E
```

## Figure 4. Student interest and experience in data science
Save composite figure

```{r}
fig4 <- plot_grid(plot_I, plot_E, labels = c("A", "B"), nrow = 2, align = "v")

fig4

ggsave(filename="Fig4.pdf", plot=fig4, width=19.05, height=14, units = "cm")
```
